Analysing the Factors Influencing the Wages of the US Workforce in 2023

### 2024-12-03 paper/paper.qmd
Final commit

---
title: "Analysing the Factors Influencing the Wages of the US Workforce in 2023"
subtitle: "Higher Education and Male Gender as Key Determinants of Wage Increases Controlling for Region and Race"
author: 
  - Jianing Li
thanks: "Code and data are available at: https://github.com/JianingLi1225/Determinants-of-Wages.git"
date: today
date-format: long
abstract: "This study uses a multiple linear regression model to analyze key factors influencing wages among the U.S. workforce, including education, age, gender, race, region, and hours worked. The results show that education is the strongest factor, with higher levels of education leading to significantly higher wages, while gender and race reveal notable disparities, with men and Asian workers earning the most. The study also finds a nonlinear relationship between age and wages, with region and hours worked playing smaller but significant roles. By identifying these drivers, the research provides evidence to inform policies aimed at reducing wage gaps and promoting fairer economic outcomes."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

The study finds that education is the strongest determinant of income. Higher levels of education lead to significantly higher earnings, with those with a college degree earning much more than other groups. The relationship between age and earnings is non-linear, with earnings peaking between the ages of 40 and 50 before declining, reflecting the impact of the occupational life cycle. In addition, gender and race show disparities, with men and Asians earning the highest average incomes, while women and certain minority groups earn less. Region and hours worked also affect earnings, but their effects are relatively small. Regional income differences reflect differences in economic development and policy environments, while hours worked show diminishing returns beyond a certain point. The findings are important because they reveal the role of education, gender, and race in driving income inequality and provide clear directions for policy interventions. They also provide empirical support for strategies to reduce income inequality, such as improving the allocation of educational resources, promoting gender equality, and increasing economic opportunities for minority groups.

@@

For this study, data from the 2023 ACS sample in IPUMS USA were used. The selected variables include STATEFIP (state code), SEX (gender), AGE (age), RACE (race), EDUC (educational attainment), EMPSTAT (employment status), UHRSWORK (usual hours worked per week), and INCWAGE (wage and salary income). These variables cover geographic location, demographic characteristics, education, and employment conditions, providing a broad basis for examining factors influencing wage income. The cleaned dataset is presented in @tbl-datasummary, along with detailed descriptions of variable definitions and construction methods, which are thoroughly explained in @sec-data.

@@

Despite its robust sampling methodology and extensive coverage, the ACS data has some inherent limitations. Variables such as income and work hours rely heavily on self-reported responses. This introduces potential inaccuracies due to recall bias, rounding errors, or deliberate misreporting. For individuals with irregular or multiple income sources, questions about wages or hours worked may be interpreted differently, further increasing measurement errors.

@@

This study use Multiple Linear Regression (MLR) to model the relationship between income and various predictors, implemented using R's `lm` function. The dataset is divided into training and testing sets using the `createDataPartition` function from the `caret` package, with 60% allocated for model training and parameter estimation and 40% for testing to evaluate predictive performance.

@@

Age also has a strong impact on income. The positive coefficient for age shows that income increases as people get older. However, the negative coefficient for age squared suggests that this growth slows over time and may decline later in life. The confidence interval for age is narrow, indicating a precise estimate. Gender differences are also evident. Male individuals have significantly higher income, as shown by the positive coefficient for gender. The narrow confidence interval for gender confirms the reliability of this effect.

Racial and ethnic factors present more complex dynamics. For example, "Black" and "Other" groups have negative coefficients and wider confidence intervals, reflecting greater uncertainty in their effects. In contrast, "White" individuals have coefficients close to zero and narrow intervals, showing minimal differences from the reference group. Regional effects are also notable. The "West" has positive coefficients compared to the reference region, while other regions show smaller or slightly negative effects. Some regional variables have confidence intervals that include zero, suggesting these effects may not be statistically significant.

@@

## Dataset Description {#sec-data}

>>>>

---
title: "Analysing the Factors Influencing the Wages of the US Workforce in 2023"
subtitle: "Higher Education and Male Gender as Key Determinants of Wage Increases Controlling for Region and Race"
author: 
  - Jianing Li
thanks: "Code and data are available at: https://github.com/JianingLi1225/Determinants-of-Wages.git"
date: today
date-format: long
abstract: "This study uses a multiple linear regression model to analyze key factors influencing wages among the U.S. workforce, including education, age, gender, race, region, and hours worked. The results show that education is the strongest factor, with bachelor’s degree holders earning nearly 50% more than those without a high school diploma, while men earn more than women, and Asian workers have the highest earnings among racial groups. The study also finds a nonlinear relationship between age and wages, with region and hours worked playing smaller but significant roles. By identifying these drivers, the research provides evidence to inform policies aimed at reducing wage gaps and promoting fairer economic outcomes."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

The study finds that education is the strongest determinant of income. Higher levels of education lead to higher earnings, with individuals holding a college degree earning nearly 50% more than those without a high school diploma. The relationship between age and earnings is non-linear, with earnings peaking between the ages of 40 and 50 before declining, reflecting the impact of the occupational life cycle. In addition, gender and race show disparities, with men and Asians earning the highest average incomes, while women and certain minority groups earn less. Region and hours worked also affect earnings, but their effects are relatively small. Regional income differences reflect differences in economic development and policy environments, while hours worked show diminishing returns beyond a certain point. The findings are important because they reveal the role of education, gender, and race in driving income inequality and provide clear directions for policy interventions. They also provide empirical support for strategies to reduce income inequality, such as improving the allocation of educational resources, promoting gender equality, and increasing economic opportunities for minority groups.

@@

For this study, data from the 2023 ACS sample in IPUMS USA were used. The selected variables include STATEFIP (state code), SEX (gender), AGE (age), RACE (race), EDUC (educational attainment), EMPSTAT (employment status), UHRSWORK (usual hours worked per week), and INCWAGE (wage and salary income). These variables cover geographic location, demographic characteristics, education, and employment conditions, providing a broad basis for examining factors influencing wage income. The cleaned dataset is presented in @tbl-datasummary, along with detailed descriptions of variable definitions and construction methods, which are thoroughly explained in @sec-description.

@@

Despite its robust sampling methodology and extensive coverage, the ACS data has some limitations. Variables such as income and work hours rely heavily on self-reported responses. This introduces potential inaccuracies due to recall bias, rounding errors, or deliberate misreporting. For individuals with irregular or multiple income sources, questions about wages or hours worked may be interpreted differently, further increasing measurement errors.

@@

This study use Multiple Linear Regression (MLR) to model the relationship between income and various predictors, carried out with the `lm` function in R [@citeR]. The dataset is divided into training and testing sets using the `createDataPartition` function from the `caret` [@caret] package, with 60% allocated for model training and parameter estimation and 40% for testing to evaluate predictive performance.

@@

Age also has a strong impact on income. The positive coefficient for age shows that income increases as people get older. However, the negative coefficient for age squared suggests that this growth slows over time and may decline later in life. The confidence interval for age is narrow, indicating a precise estimate. Gender differences are also evident. Male individuals have higher income, as shown by the positive coefficient for gender. The narrow confidence interval for gender confirms the reliability of this effect.

Racial factor present more complex dynamics. For example, "Black" and "Other" groups have negative coefficients and wider confidence intervals, reflecting greater uncertainty in their effects. In contrast, "White" individuals have coefficients close to zero and narrow intervals, showing minimal differences from the reference group. Regional effects are also notable. The "West" has positive coefficients compared to the reference region, while other regions show smaller or slightly negative effects. Some regional variables have confidence intervals that include zero, suggesting these effects may not be statistically significant.

@@

## Dataset Description {#sec-description}
